---
title: "Trees in R Assignment"
output:
  html_document:
      toc: true
      toc_depth: 2
      theme:
        bootswatch: cerulean
---

## Abstract

The tree being investigated here is one of all extant pinnipeds alongside 2 outgroups. The branch lengths in the tree is based on were calculated

## Tree Metrics

#### Set up

###### Loading required packages:

```{r load in packages, results='hide', message=FALSE}
library(ape)
library(bslib) #allows access to bootswatch themes for markdown
```

###### Loading in the tree:

The raw data for this tree was downloaded from the Open Tree of Life datastore, (McTavish EJ, et. al., 2015), and can be accessed via the following link <https://tree.opentreeoflife.org/curator/study/view/pg_1646/?tab=home>.

```{r reading in tree from newick data}
tree <- read.tree(text = "((((((((((((((Arctocephalus_australis:0.8,Arctocephalus_galapagoensis:0.8):0.3,Arctocephalus_forsteri:1.1):1.9,(Arctocephalus_gazella:0.1,Arctocephalus_tropicalis:0.1):2.9):0.1,(Arctocephalus_philippii:0.3,Arctocephalus_townsendi:0.3):2.8):0.2,Phocarctos_hookeri:3.3):0.9,Arctocephalus_pusillus:4.2):0.2,Otaria_byronia:4.4):0.8,(Eumetopias_jubatus:4.5,Zalophus_californianus:4.5):0.7):0.4,Neophoca_cinerea:5.6):2.6,Callorhinus_ursinus:8.2):9.8,Odobenus_rosmarus:18):5,(((Cystophora_cristata:7.9,(((((Halichoerus_grypus:1.7,Pusa_caspica:1.7):0.1,Pusa_sibirica:1.8):0.1,(Phoca_largha:0.9,Phoca_vitulina:0.9):1):0.2,Pusa_hispida:2.1):4.3,(Histriophoca_fasciata:4.3,Pagophilus_groenlandica:4.3):2.1):1.5):5.1,Erignathus_barbatus:13):3,(((((Hydrurga_leptonyx:4.3,Leptonychotes_weddellii:4.3):2.5,Ommatophoca_rossii:6.8):0.3,Lobodon_carcinophagus:7.1):2.8,(Mirounga_angustirostris:2.3,Mirounga_leonina:2.3):7.6):1.4,(Monachus_monachus:9.9,(Monachus_schauinslandi:4.9,Monachus_tropicalis:4.9):5):1.4):4.7):7):12.7,Ursidae:35.7):7.7,Canidae:43.4);")
```

###### Plotting Tree:

```{r plotting tree, fig.align='center', fig.height= 9, fig.width= 9, fig.cap="Fig 1: Phylogenetic tree of the Pinnipeds with axis in millions of years based on data from (Hidgon, et. al., 2007)"}
ladderised_tree <- ladderize(tree, right=FALSE) #Used to make it easier to see how the time scale maps onto the tree.
plot(ladderised_tree, cex = 0.7)
axisPhylo(cex = 0.5)
```

This tree shows the 33 extant species of pinnipeds, as well as the now extinct Carribean Monk Seal (*Monachus tropicalis*) in addition to Canidae and Ursidae as the outgroups. *Mustelidae* may have been a more appropriate outgroup as since this study it has been shown that they are the sister family to *Pinnepedia* (Park, et. al., 2024). It also shows Odobenids sitting within the *Otariidae* which is the better supported hypothesis, however it was once thought they might group within *Phocidae* instead (Hidgon, et. al, 2007)

#### Analysis

###### Tree Topology:

Although these features can often be inferred from looking at the shape of the tree I decided to test for both whether the tree was ultrametric and bifurcating.

```{r testing if tree is ultrametric and bifurcating}
is.ultrametric(tree)
is.binary(tree)
```

The tree being ultrametric was expected as this is a tree of all extant pinnipeds and both of the outgroups used are also clades that still exist today so all of the branches should continue until the present day.

Although it does match the observed branching pattern of the tree it was slightly unexpected that the tree was found to be bifurcating. This is because this suggeststhe tree contains no polytomies. This is interesting however as it has been suggested that polytomies could exist both at the basal end of the tribe *Phocini* (one of three tribes that make up the *Phocinae* subfamily of earless/true seals) and potentially within the genus *Pusa* (Hidgon, et. al., 2007).

It is also important to check these as some of the functions used to analyse tree metrics assume that trees are ultrametric and bifurcating.

###### Phylogenetic diversity:

```{r Whole tree phylogenetic diversity}
sum(tree$edge.length)
```

As the branch lengths in this tree are based on divergence dates in millions of years the total sum of branch lengths can be used to indicate the total amount of evolutionary history which is contained within this tree. This shows that this tree contains 319.8 years of evolutionary history, however this also includes the *Canidae* and *Ursidae* out groups.

From this point forward the tree containing no outgroups will be used so that analyses are concerning solely the pinnipeds.

```{r Creating tree without outgroups, echo=FALSE}
no_outgroup_tree <- drop.tip(tree, c("Canidae", "Ursidae"))
```

```{r Pinniped phylogenetic diversity calculation, echo=FALSE}
phylo_div <- sum(no_outgroup_tree$edge.length)
```

```{r Phylogenetic Diversity}
phylo_div
```


Running this without the outgroups included shows that from the last common pinniped ancestor there is 220.3 million years of evolutionary history conserved.

###### Tree Imbalance:

```{r Function for calculating imbalance metric, echo=FALSE}
imbalance_metric <-function (tr) {
x<-balance(tr)
for (i in (1:nrow(x))) {
if (x[i,1]<x[i,2]) {
x[i,1:2]<- x[i,2:1]}
}
return(mean(x[,1]/(x[,1]+x[,2])))}
```

```{r Pinniped Imbalance Metric Calculation, echo=FALSE}
imbalance <- imbalance_metric(no_outgroup_tree)
```

```{r Imbalance Metric}
imbalance
```


Imbalance metric is 0.70, showing that while the tree is not completely balanced it is still overall more balanced than unbalanced. It also suggests there is a slight tendency for progressive branching, which can be seen specifically in the genus *Arctocephalus* and also the tribe *Phocina* (includes the genus *Phoca* as well as *Halicheoeris grypus*).

###### Speciation Rate:

Speciation rate can calculated by fitting the yule process to a tree which will allow you to calculate the instantaneous speciation rate λ, which can be multiplied by the number of species to calculate the total speciation rate (Crawford & Suchard, 2013)

```{r Speciation Rate Calculation, echo=FALSE}
yule_result <- yule(no_outgroup_tree)
speciation_rate <- yule_result$lambda* (no_outgroup_tree$Nnode + 1) #Nnode +1 represents the number of species
```

```{r Speciation Rate}
speciation_rate
```


## Metrics Table:

```{r Setting up Table, echo=FALSE}
tree_metrics_matrix <- matrix(c(phylo_div, imbalance, speciation_rate), ncol = 3, byrow = TRUE, nrow = 1)

colnames(tree_metrics_matrix) <- c('Phylogenetic Diversity','Imbalance','Speciation Rate')
rownames(tree_metrics_matrix) <- "Values"

metric_table <- as.table(tree_metrics_matrix)

```

```{r Table, echo=FALSE}
knitr::kable(tree_metrics_matrix)
```


## Bonus Analysis - Comparing metrics within clades

Although there are technically three extant families within the Pinnipeds as there is only one living species within *Odobenidae* therefore I will just be comparing *Otariidae* and *Phocidae* in this analysis.

<!--I also wanted to see if the two main families, Otariidae and Phocidae where equally diverse, however I did not want this code to interfere with the ability of the main tree analysis to automatically reproduce figures and tables if topology if the original tree is changed, hence why this section has been moved to the end of the report-->

Otariidae imbalance:

```{r Creating Otariidae Sub-Tree, echo=FALSE}
subtrees <- subtrees(tree)
otariidae_subtree <- subtrees[[5]]
```

```{r Otariidae Imbalance }
imbalance_metric(otariidae_subtree)
```

Phocidae Imbalance:

```{r Creating Phocidae Sub-Tree, echo=FALSE}
subtrees <- subtrees(tree)
phocidae_subtree <- subtrees[[18]]
```

```{r Phocidae Imbalance}
imbalance_metric(phocidae_subtree)
```

```{r Function for creating a balance table,  echo=FALSE}
imbalance_table <-function (tr) {
x<-balance(tr)
for (i in (1:nrow(x))) {
if (x[i,1]<x[i,2]) {
x[i,1:2]<- x[i,2:1]}
}
return(data.frame(x[,1]/(x[,1]+x[,2])))}
```

```{r Otariidae Imbalance Table, echo=FALSE, results='hide'}
otariidae_imbalance_table <-(imbalance_table(otariidae_subtree))
colnames(otariidae_imbalance_table)[1] ="Balance"
otariidae_imbalance_table
```

```{r Phocidae Imbalance Table, echo=FALSE, results='hide'}
phocidae_imbalance_table <-(imbalance_table(phocidae_subtree))
colnames(phocidae_imbalance_table)[1] ="Balance"
phocidae_imbalance_table
```

F test of equal variance:

```{r F Test}
var.test(phocidae_imbalance_table$Balance, otariidae_imbalance_table$Balance)
```

P value is 0.3054, as this is \> 0.05 we do not reject the null hypothesis and can conclude that the variances a likely not significantly different.

```{r 2 Sample T Test}
t.test(phocidae_imbalance_table$Balance, otariidae_imbalance_table$Balance, var.equal = TRUE)
```

P value is 0.2057 meaning that we do not reject the null hypothesis and conclude that the imbalance metric between Phocidae and Otariidae is not statistically significant.

I also decided to investigate whether the two families are equally diverse and contain similar amounts of evolutionary history.

```{r Otariid Diversity}
sum(otariidae_subtree$edge.length)
```

```{r Phocidae Diversity}
sum(phocidae_subtree$edge.length)
```
